As discussed previously, capacitance, $C$, is defined as
	\begin{align*}
		U_C    &=   \frac{1}{2C}q^2   . \\
	\intertext{Similarly, inductance, $L$, is defined as}
		U_L    &=    \frac{L}{2}I^2,
	\end{align*}
	and measured in units of  henries.
	The magnitude of the voltage difference across a capacitor or inductor is given by
	\begin{align*}
		V_C	&= \frac{q}{C} \\
	\intertext{or}
		V_L	&= L\frac{\der I}{\der t}.
	\end{align*}
	In the equation for the inductor, the
	direction of the voltage drop (plus or minus sign) is such
	that the inductor resists the change in current.
	
	A series LRC circuit exhibits \emph{oscillation}, and, if driven
	by an external voltage, resonates.
	The $Q$ of the circuit relates to the resistance value. For large
	$Q$, the resonant frequency is
	\begin{equation*}
	  \omega \approx \frac{1}{\sqrt{LC}}.
	\end{equation*}
	A series RC or RL circuit exhibits exponential \emph{decay},
	\begin{align*}
		q &= q_\zu{o}\exp\left(-\frac{t}{RC}\right) \\
	\intertext{or}
		I &= I_\zu{o}\exp\left(-\frac{R}{L}t\right),
	\end{align*}
	and the quantity $RC$ or $L/R$ is known as the time constant.

	When driven by a sinusoidal AC voltage with amplitude $\tilde{V}$,
	a capacitor, resistor, or inductor responds with a current
	having amplitude
	\begin{equation*}
		\tilde{I} = \frac{\tilde{V}}{Z}  ,
	\end{equation*}
	where the \emph{impedance\/}, $Z$, is a frequency-dependent quantity
	having units of ohms.
	In a capacitor, the current has a phase that is $90\degunit$ ahead of the voltage,
	while in an inductor the current is $90\degunit$ behind.
        We can represent these phase relationships by defining the impedances
	 as complex numbers:
	 \begin{align*}
		Z_C &= -\frac{i}{\omega C}\\
		Z_R	&= R\\
		Z_L &= i\omega L 
	\end{align*}
	The arguments of the complex impedances are to be interpreted as phase relationships between the
	oscillating voltages and currents.
	The complex impedances defined in this way combine in series and parallel
	 according to the same rules as resistances.

	When a voltage source is driving a load through a transmission line, the maximum
	power is delivered to the load when the impedances of the line and the load
	are matched.
