<% begin_sec("Some theorems and proofs",nil,'am-theorems') %>
In this section I prove three theorems stated earlier,
and state a fourth theorem whose proof is left as an exercise.

<% begin_sec("Uniqueness of the cross product") %>\label{misc:uniquexproof}
The vector cross product as we have defined it has the
following properties:\\
(1) It does not violate rotational invariance.\\
(2) It has the property $\zb{A}\times(\zb{B}+\zb{C})=\zb{A}\times\zb{B}+\zb{A}\times\zb{C}$.\\
(3) It has the property $\zb{A}\times(k\zb{B})=k(\zb{A}\times\zb{B})$, where $k$ is a scalar.


\mythmhdr{Theorem} The definition we have given is the only possible method of
multiplying two vectors to make a third vector which has
these properties, with the exception of trivial
redefinitions which just involve multiplying all the results
by the same constant or swapping the names of the axes.
(Specifically, using a left-hand rule rather than a
right-hand rule corresponds to multiplying all the results
by $-1$.)

\mythmhdr{Proof} We prove only the uniqueness of the definition, without
explicitly proving that it has properties (1) through (3).

Using properties (2) and (3), we can break down any vector
multiplication
$(A_x\hat{\zb{x}}+A_y\hat{\zb{y}}+A_z\hat{\zb{z}})
\times(B_x\hat{\zb{x}}+B_y\hat{\zb{y}}+B_z\hat{\zb{z}})$ into terms involving
cross products of unit vectors.

A ``self-term'' like $\hat{\zb{x}}\times\hat{\zb{x}}$ must either be zero or lie along the $x$
axis, since any other direction would violate property (1).
If it was not zero, then
$(-\hat{\zb{x}})\times(-\hat{\zb{x}})$
 would have to lie in the opposite direction to avoid
breaking rotational invariance, but property (3) says that 
$(-\hat{\zb{x}})\times(-\hat{\zb{x}})$ is the
same as $\hat{\zb{x}}\times\hat{\zb{x}}$, which is a contradiction.
Therefore the self-terms
must be zero.

An ``other-term'' like $\hat{\zb{x}}\times\hat{\zb{y}}$ could conceivably have components in
the $x$-$y$ plane and along the $z$ axis. If there was a nonzero
component in the $x$-$y$ plane, symmetry would require that it
lie along the diagonal between the $x$ and $y$ axes, and
similarly the in-the-plane component of $(-\hat{\zb{x}})\times\hat{\zb{y}}$  would have to
be along the other diagonal in the $x$-$y$ plane. Property (3),
however, requires that $(-\hat{\zb{x}})\times\hat{\zb{y}}$ equal
$-(\hat{\zb{x}}\times\hat{\zb{y}})$, which would be along
the original diagonal. The only way it can lie along both
diagonals is if it is zero.

We now know that $\hat{\zb{x}}\times\hat{\zb{y}}$
 must lie along the $z$ axis. Since we are
not interested in trivial differences in definitions, we can
fix $\hat{\zb{x}}\times\hat{\zb{y}}=\hat{\zb{z}}$,
ignoring peurile possibilities such as
$\hat{\zb{x}}\times\hat{\zb{y}}=7\hat{\zb{z}}$ or the
left-handed definition $\hat{\zb{x}}\times\hat{\zb{y}}=-\hat{\zb{z}}$.
Given $\hat{\zb{x}}\times\hat{\zb{y}}=\hat{\zb{z}}$, the symmetry of space
requires that similar relations hold for $\hat{\zb{y}}\times\hat{\zb{z}}$ and
$\hat{\zb{z}}\times\hat{\zb{x}}$, with at most
a difference in sign. A difference in sign could always be
eliminated by swapping the names of some of the axes, so
ignoring possible trivial differences in definitions we can
assume that the cyclically related set of relations  $\hat{\zb{x}}\times\hat{\zb{y}}=\hat{\zb{z}}$,
 $\hat{\zb{y}}\times\hat{\zb{z}}=\hat{\zb{x}}$, and  $\hat{\zb{z}}\times\hat{\zb{x}}=\hat{\zb{y}}$
holds. Since the arbitrary cross-product with which we
started can be broken down into these simpler ones, the
cross product is uniquely defined.
<% end_sec %>

<% begin_sec("Choice of axis theorem") %>
\index{choice of axis theorem!proof}\index{angular momentum!choice of axis theorem}
\mythmhdr{Theorem} Suppose a closed system of material particles conserves
angular momentum in one frame of reference, with the axis taken to be
at the origin. Then conservation of angular momentum is unaffected if the origin
is relocated or if we change to a frame of reference that is in constant-velocity
motion with respect to the first one. The theorem also holds in the case where the system is not
closed, but the total external force is zero.\label{choiceofaxisproof}

\mythmhdr{Proof} In the original frame of reference, angular momentum is
conserved, so we have $\der\zb{L}/\der t$=0.
From example \ref{eg:torqueproof} on page \pageref{eg:torqueproof}, this
derivative can be rewritten as
\begin{equation*}
		\frac{\der\zb{L}}{\der t}	= \sum_i  \zb{r}_i\times\zb{F}_i\eqquad,
\end{equation*}
where $\zb{F}_i$ is the total force acting on particle $i$. In other words, we're
adding up all the torques on all the particles. 

By changing to the new frame of reference, we have changed
the position vector of each particle according to $\zb{r}_i \rightarrow \zb{r}_i+\zb{k}-\zb{u}t$,
where \zb{k} is a constant vector that indicates the relative position of the new
origin at $t=0$, and \zb{u} is the velocity of the new frame with respect to the old one.
The forces are all the same in the new frame of reference, however.
In the new frame, the rate of change of the angular momentum is
\begin{align*}
		\frac{\der\zb{L}}{\der t}	&= \sum_i
				(\zb{r}_i+\zb{k}-\zb{u}t)\times\zb{F}_i \\
		&= \sum_i \zb{r}_i\times\zb{F}_i
			+ (\zb{k}-\zb{u}t) \times \sum_i \zb{F}_i\eqquad.
\end{align*}
The first term is the expression for the rate of change of the angular
momentum in the original frame of reference, which is zero by
assumption. The second term vanishes by Newton's third law; since the system is
closed, every force $\zb{F}_i$ cancels with some force $\zb{F}_j$.
(If external forces act, but they add up to zero, then the sum can be
broken up into a sum of internal forces and a sum of external forces, each of
which is zero.)
The rate of change of the angular momentum is therefore zero
in the new frame of reference.
<% end_sec %>

<% begin_sec("Spin theorem") %>
\index{angular momentum!spin theorem}\index{spin theorem!proof}
Theorem: An object's angular momentum with respect to some outside
axis A can be found by adding up two parts:\\
(1) The first part is the object's angular momentum
found by using its own center of mass as the axis, i.e., the
angular momentum the object has because it is spinning.\\
(2) The other part equals the angular momentum that the object
would have with respect to the axis A if it had all its mass
concentrated at and moving with its center of mass.

Proof: Let the system's center of mass be at  $\vc{r}_{cm}$, and let particle $i$ lie at position $\vc{r}_{cm}+\vc{d}_i$.
Then the total angular momentum is
\begin{align*}
  \vc{L} &= \sum_i (\vc{r}_{cm}+\vc{d}_i)\times\vc{p}_i \\
         &= \vc{r}_{cm}\times\sum_i \vc{p}_i+\sum_i \vc{d}_i\times\vc{p}_i\eqquad,
\end{align*}
which establishes the result claimed, since we can identify the first term with (2)
and the second with (1).
<% end_sec %>

<% begin_sec("Parallel axis theorem") %>\label{parallel-axis-theorem-statement}\index{parallel axis theorem}
Suppose an object has mass $m$, and moment of inertia $I_\zu{o}$ for rotation about some axis A
passing through its center of mass. Given a new axis B, parallel to A and lying at a distance $h$ from
it, the object's moment of inertia is given by $I_\zu{o}+mh^2$.

The proof of this theorem is left as an exercise (problem \ref{hw:parallel-axis-theorem}, p.~\pageref{hw:parallel-axis-theorem}).
<% end_sec %>

<% end_sec %>
