<% begin_sec("Proof of Kepler's elliptical orbit law",nil,'elliptical-orbits',{'optional'=>true}) %>%
\index{Kepler!elliptical orbit law}\index{elliptical orbit law}

Kepler determined purely empirically that the planets'
orbits were ellipses, without understanding the underlying
reason in terms of physical law. Newton's proof of this fact
based on his laws of motion and law of gravity was
considered his crowning achievement both by him and by his
contemporaries, because it showed that the same physical
laws could be used to analyze both the heavens and the
earth. Newton's proof was very lengthy, but by applying the
more recent concepts of conservation of energy and angular
momentum we can carry out the proof quite simply and
succinctly, and without calculus.

The basic idea of the proof is that we want to describe the
shape of the planet's orbit with an equation, and then show
that this equation is exactly the one that represents an
ellipse. Newton's original proof had to be very complicated
because it was based directly on his laws of motion, which
include time as a variable. To make any statement about the
shape of the orbit, he had to eliminate time from his
equations, leaving only space variables. But conservation
laws tell us that certain things don't change over time, so
they have already had time eliminated from them.

<% marg(50) %>
<%
  fig(
    'proof-a',
    %q{The $r-\varphi$ representation of a curve.}
  )
%>
\spacebetweenfigs
<%
  fig(
    'proof-b',
    %q{%
      Proof that the two angles labeled 
      $\varphi$ are in fact equal: The definition of an ellipse is that
      the sum of the distances from the two foci stays constant. If we move a small distance
      $\ell$ along the ellipse, then one distance shrinks by an amount 
      $\ell\cos\varphi_1$, while the other
      grows by $\ell\cos\varphi_2$. These are equal, so $\varphi_1=\varphi_2$.%
      
    }
  )
%>

<% end_marg %>
There are many ways of representing a curve by an equation,
of which the most familiar is $y=ax+b$ for a line in
two dimensions. It would be perfectly possible to describe a
planet's orbit using an $x-y$ equation like this, but
remember that we are applying conservation of angular
momentum, and the space variables that occur in the equation
for angular momentum are the distance from the axis, $r$,
and the angle between the velocity vector and the $r$
vector, which we will call $\varphi $. The planet will have
$\varphi $=90\degunit when it is moving perpendicular to the $r$
vector, i.e., at the moments when it is at its smallest or
greatest distances from the sun. When $\varphi $ is less than
90\degunit the planet is approaching the sun, and when it is
greater than 90\degunit it is receding from it. Describing a
curve with an $r-\varphi $ equation is like telling a driver in
a parking lot a certain rule for what direction to steer
based on the distance from a certain streetlight in
the middle of the lot.

The proof is broken into the three parts for easier
digestion. The first part is a simple and intuitively
reasonable geometrical fact about ellipses, whose proof we
relegate to the caption of figure \figref{proof-b}; you will not be missing
much if you merely absorb the result without reading the proof.

(1) If we use one of the two foci of an ellipse as an axis
for defining the variables $r$ and $\varphi $, then the angle
between the tangent line and the line drawn to the other
focus is the same as $\varphi $, i.e., the two angles labeled
$\varphi $ in figure \figref{proof-b} are in fact equal.

The other two parts form the meat of our proof. We state the
results first and then prove them.

(2) A planet, moving under the influence of the sun's
gravity with less than the energy required to escape, obeys
an equation of the form
\begin{equation*}
 \sin\varphi = \frac{1}{\sqrt{-pr^2+qr}}\eqquad,
\end{equation*}
 where $p$ and $q$  are positive constants that depend on the planet's
 energy and angular momentum.

(3) A curve is an ellipse if and only if its $r-\varphi $
equation is of the form
\begin{equation*}
 \sin\varphi = \frac{1}{\sqrt{-pr^2+qr}}\eqquad,
\end{equation*}
where $p$ and $q$ are positive constants that depend on the size and
shape of the ellipse.

<% begin_sec("Proof of part (2)",0,'',{'toc'=>false}) %>
The component of the planet's velocity vector that is
perpendicular to the $\vc{r}$ vector is $v_\perp=v \sin \varphi $, so
conservation of angular momentum tells us that $L=mrv \sin\varphi$
is a constant. Since the planet's
mass is a constant, this is the same as the condition
\begin{equation*}
                rv \sin  \varphi  =  \text{constant}\eqquad.
\end{equation*}
Conservation of energy gives
\begin{equation*}
                \frac{1}{2}mv^2-\frac{GMm}{r}          =  \text{constant}\eqquad.
\end{equation*}
We solve the first equation for $v$ and plug into the second
equation to eliminate $v$. Straightforward algebra then
leads to the equation claimed above, with the constant $p$
being positive because of our assumption that the planet's
energy is insufficient to escape from the sun, i.e., its
total energy is negative.

<% marg(0) %>
<%
  fig(
    'proof-c',
    %q{Proof of part (3).}
  )
%>
<% end_marg %>
<% end_sec() %>
<% begin_sec("Proof of part (3)",0,'',{'toc'=>false}) %>
We define the quantities $\alpha$, $d$, and $s$ as shown in
the figure. The law of cosines gives
\begin{equation*}
 d^2 = r^2+s^2-2rs\cos\alpha\eqquad.
\end{equation*}
Using $\alpha =180\degunit-2\varphi$ and the trigonometric
identities $\cos (180\degunit-x)=-\cos x$ and
$\cos 2x= 1-2 \sin^2 x$, we can rewrite this as
\begin{equation*}
       d^2 = r^2 + s^2 - 2rs\left(2\sin^2\varphi-1\right)\eqquad.
\end{equation*}
Straightforward algebra transforms this into
\begin{equation*}
  \sin\:\varphi = \sqrt{\frac{(r+s)^2-d^2}{4rs}}\eqquad.
\end{equation*}
Since $r+s$ is constant, the top of the fraction is
constant, and the denominator can be rewritten as $4rs=4r(\text{constant}-r)$,
which is equivalent to the desired form.

<% end_sec() %>
<% end_sec() %>
