<% begin_sec("Forces have no perpendicular effects",0) %>

Suppose you could shoot a rifle and arrange for a second
bullet to be dropped from the same height at the exact
moment when the first left the barrel. Which would hit the
ground first? Nearly everyone expects that the dropped
bullet will reach the dirt first, and Aristotle would have
agreed. Aristotle would have described it like this. The
shot bullet receives some forced motion from the gun. It
travels forward for a split second, slowing down rapidly
because there is no longer any force to make it continue in
motion. Once it is done with its forced motion, it changes
to natural motion, i.e. falling straight down. While the
shot bullet is slowing down, the dropped bullet gets on with
the business of falling, so according to Aristotle  it will
hit the ground first.

<%
  fig(
    'bullets',
    %q{%
      A bullet is shot from a gun,
      and another bullet is simultaneously dropped from the same
      height. 1. Aristotelian physics says that the horizontal motion
      of the shot bullet delays the onset of falling, so the dropped
      bullet hits the ground first. 2. Newtonian physics says the
      two bullets have the same vertical motion, regardless of
      their different horizontal motions.
    },
    {
      'width'=>'fullpage'
    }
  )
%>

Luckily, nature isn't as complicated as Aristotle thought!
To convince yourself that Aristotle's ideas were wrong and
needlessly complex, stand up now and try this experiment.
Take your keys out of your pocket, and begin walking briskly
forward. Without speeding up or slowing down, release your
keys and let them fall while you continue walking at the same pace.

\enlargethispage{-\baselineskip}

You have found that your keys hit the ground right next to
your feet. Their horizontal motion never slowed down at all,
and the whole time they were dropping, they were right next
to you. The horizontal motion and the vertical motion happen
at the same time, and they are independent of each other.
Your experiment proves that the horizontal motion is
unaffected by the vertical motion, but it's also true that
the vertical motion is not changed in any way by the
horizontal motion. The keys take exactly the same amount of
time to get to the ground as they would have if you simply
dropped them, and the same is true of the bullets: both
bullets hit the ground simultaneously.

These have been our first examples of motion in more than
one dimension, and they illustrate the most important new
idea that is required to understand the three-dimensional
generalization of Newtonian physics:

\begin{important}[Forces have no perpendicular effects.]
When a force acts on an object, it has no effect on the part
of the object's motion that is perpendicular to the force.
\end{important}

\noindent In the examples above, the vertical force of gravity had no
effect on the horizontal motions of the objects. These were
examples of projectile motion, which interested people like
Galileo because of its military applications. The principle
is more general than that, however. For instance, if a
rolling ball is initially heading straight for a wall, but a
steady wind begins blowing from the side, the ball does not
take any longer to get to the wall. In the case of
projectile motion, the force involved is gravity, so we can
say more specifically that the vertical acceleration is 9.8
$\munit/\sunit^2$, regardless of the horizontal motion.

<% self_check('blown-sideways',<<-'SELF_CHECK'
In the example of the ball being blown sideways, why doesn't
the ball take longer to get there, since it has to
travel a greater distance?
  SELF_CHECK
  ) %>

\enlargethispage{-\baselineskip}


<% begin_sec("Relationship to relative motion") %>

These concepts are directly related to the idea that motion
is relative. Galileo's opponents argued that the earth could
not possibly be rotating as he claimed, because then if you
jumped straight up in the air you wouldn't be able to come
down in the same place. Their argument was based on their
incorrect Aristotelian assumption that once the force of
gravity began to act on you and bring you back down, your
horizontal motion would stop. In the correct Newtonian
theory, the earth's downward gravitational force is acting
before, during, and after your jump, but has no effect on
your motion in the perpendicular (horizontal) direction.

If Aristotle had been correct, then we would have a handy
way to determine absolute motion and absolute rest: jump
straight up in the air, and if you land back where you
started, the surface from which you jumped must have been in
a state of rest. In reality, this test gives the same result
as long as the surface under you is an inertial frame. If
you try this in a jet plane, you land back on the same spot
on the deck from which you started, regardless of whether
the plane is flying at 500 miles per hour or parked on the
runway. The method would in fact only be good for detecting
whether the plane was accelerating.

\startdqs

\begin{dq}
The following is an incorrect explanation of a fact
about target shooting:

``Shooting a high-powered rifle with a high muzzle velocity
is different from shooting a less powerful gun. With a less
powerful gun, you have to aim quite a bit above your target,
but with a more powerful one you don't have to aim so high
because the bullet doesn't drop as fast.''

Explain why it's incorrect. What is the correct explanation?
\end{dq}

<%
  fig(
    'dq-target-shooting',
    '',
    {
      'width'=>'wide',
      'anonymous'=>true,
      'float'=>false
    }
  )
%>

\begin{dq}
You have thrown a rock, and it is flying through the air
in an arc.  If the earth's gravitational force on it is
always straight down, why doesn't it just go straight down
once it leaves your hand?
\end{dq}

\begin{dq}
Consider the example of the bullet that is dropped at the
same moment another bullet is fired from a gun. What would
the motion of the two bullets look like to a jet pilot
flying alongside in the same direction as the shot bullet
and at the same horizontal speed?
\end{dq}

<%
  fig(
    'projectile-on-grid',
    %q{%
      This object experiences a force that pulls it
      down toward the bottom of the page. In each
      equal time interval, it moves three units to
      the right. At the same time, its vertical motion is
       making a simple pattern of $+1$, 0, $-1$,
      $-2$, $-3$, $-4$, \ldots units. Its motion can be 
      described by an $x$ coordinate that has
       zero acceleration and a $y$ coordinate with constant
      acceleration. The arrows labeled $x$ and $y$
      serve to explain that we are defining increasing
      $x$ to the right and increasing $y$ as upward.
    },
    {
      'width'=>'wide',
      'sidecaption'=>false
    }
  )
%>

<% end_sec() %>
<% end_sec() %>
<% begin_sec("Coordinates and components",0) %> % no page break because of figure above heading

\epigraphlong{'Cause we're all\\
Bold as love,\\
Just ask the axis.}{Jimi Hendrix}

How do we convert these ideas into mathematics? Figure \figref{projectile-on-grid}
shows a good way of connecting the intuitive ideas to
the numbers. In one dimension, we impose a number line with
an $x$ coordinate on a certain stretch of space. In two
dimensions, we imagine a grid of squares which we label with
$x$ and $y$ values, as shown in figure \figref{projectile-on-grid}.

<% marg(55) %>
<%
  fig(
    'shadows',
    %q{%
      The shadow on the wall shows the ball's $y$
      motion, the shadow on the floor its $x$ motion.
    }
  )
%>
<% end_marg %>
But of course motion doesn't really occur in a series of
discrete hops like in chess or checkers. Figure \figref{shadows}
shows a way of conceptualizing the smooth variation of
the $x$ and $y$ coordinates. The ball's shadow on the wall
moves along a line, and we describe its position with a
single coordinate, $y$, its height above the floor. The wall
shadow has a constant acceleration of -9.8 $\munit/\sunit^2$. A shadow
on the floor, made by a second light source, also moves
along a line, and we describe its motion with an $x$
coordinate, measured from the wall.

The velocity of the floor shadow is referred to as the
$x$ component of\index{component!defined} the velocity,
written $v_x$. Similarly we can notate the acceleration of
the floor shadow as $a_x$. Since $v_x$ is constant, $a_x$ is zero.

Similarly, the velocity of the wall shadow is called $v_y$,
its acceleration $a_y$. This example has $a_y=-9.8\ \munit/\sunit^2$.

Because the earth's gravitational force on the ball is
acting along the $y$ axis, we say that the force has a
negative $y$ component, $F_y$, but $F_x=F_z=0$.

The general idea is that we imagine two observers, each of
whom perceives the entire universe as if it was flattened
down to a single line. The $y$-observer, for instance,
perceives $y$, $v_y$, and $a_y$, and will infer that there
is a force, $F_y$, acting downward on the ball. That is, a
$y$ component means the aspect of a physical phenomenon,
such as velocity, acceleration, or force, that is observable
to someone who can only see motion along the $y$ axis.

All of this can easily be generalized to three dimensions.
In the example above, there could be a $z$-observer who only
sees motion toward or away from the back wall of the room.

\pagebreak[4]

<% marg(0) %>
<%
  fig(
    'eg-car-crash',
    %q{Example \ref{eg:car-crash}.}
  )
%>
<% end_marg %>
\begin{eg}{A car going over a cliff}\label{eg:car-crash}
\egquestion The police find a car at a distance $w=20\ \munit$ from
the base of a cliff of height $h=100\ \munit$. How fast was the car
going when it went over the edge? Solve the problem
symbolically first, then plug in the numbers.

\eganswer Let's choose $y$ pointing up and $x$ pointing away
from the cliff. The car's vertical motion was independent of
its horizontal motion, so we know it had a constant vertical
acceleration of $a=-g=-9.8\ \munit/s^2$. The time it spent in
the air is therefore related to the vertical distance it
fell by the constant-acceleration equation
\begin{align*}
  \Delta y &= \frac{1}{2}a_y \Delta t^2\eqquad, \\
\intertext{or}
  -h &= \frac{1}{2}(-g)\Delta t^2\eqquad.
\end{align*}
Solving for $\Delta t$ gives
\begin{equation*}
  \Delta t = \sqrt{\frac{2h}{g}}\eqquad.
\end{equation*}
Since the vertical force had no effect on the car's
horizontal motion, it had $a_x=0$, i.e., constant horizontal
velocity. We can apply the constant-velocity equation
\begin{align*}
       v_x &= \frac{\Delta x}{\Delta t}\eqquad, \\
\intertext{i.e.,}
       v_x &= \frac{w}{\Delta t}\eqquad.
\end{align*}
We now substitute for $\Delta t$ to find
\begin{equation*}
      v_x = w / \sqrt{\frac{2h}{g}}\eqquad,
\end{equation*}
which simplifies to
\begin{equation*}
      v_x = w \sqrt{\frac{g}{2h}}\eqquad.
\end{equation*}
Plugging in numbers, we find that the car's speed when it
went over the edge was 4 m/s, or about 10 mi/hr.
\end{eg}

\index{projectiles}\index{parabola!motion of projectile on}
<% begin_sec("Projectiles move along parabolas.") %>

What type of mathematical curve does a projectile follow
through space? To find out, we must relate $x$ to $y$,
eliminating $t$. The reasoning is very similar to that used
in the example above. Arbitrarily choosing $x=y=t=0$ to be
at the top of the arc, we conveniently have 
$x=\Delta x$, $y=\Delta y$, and $t=\Delta t$, so
\begin{align*}
  y &= \frac{1}{2}a_y t^2 \qquad (a_y<0)\\
  x &= v_x t
\end{align*}
We solve the second equation for $t=x/v_x$  and eliminate
$t$ in the first equation:
\begin{equation*}
  y = \frac{1}{2}a_y\left(\frac{x}{v_x}\right)^2 .
\end{equation*}
Since everything in this equation is a constant except for
$x$ and $y$, we conclude that $y$ is proportional to the
square of $x$. As you may or may not recall from a math
class, $y\propto x^2$ describes a parabola.

<% marg(75) %>
<%
  fig(
    'parabola',
    %q{%
      A parabola can be defined as the
      shape made by cutting a cone parallel to its side. A parabola is also the
      graph of an equation of the form
      $y\propto x^2$.
    }
  )
%>
\spacebetweenfigs
<%
  fig(
    'hose',
    %q{%
      Each water droplet follows a parabola. The faster drops' parabolas
      are bigger.
    }
  )
%>
<% end_marg %>

m4_ifelse(__me,1,[::],\worked{cannon-range}{A cannon})

\startdq

\begin{dq}
At the beginning of this section I represented the motion
of a projectile on graph paper, breaking its motion into
equal time intervals. Suppose instead that there is no force
on the object at all. It obeys Newton's first law and
continues without changing its state of motion. What would
the corresponding graph-paper diagram look like? If the time
interval represented by each arrow was 1 second, how would
you relate the graph-paper diagram to the velocity
components $v_x$ and $v_y?$
\end{dq}

\begin{dq}
Make up several different coordinate systems oriented in
different ways, and describe the $a_x$ and $a_y$ of a
falling object in each one.
\end{dq}


<% end_sec() %>
<% end_sec() %>
<% begin_sec("Newton's laws in three dimensions",0) %>\index{Newton's laws of motion!in three dimensions}

It is now fairly straightforward to extend Newton's laws
to three dimensions:

\begin{lessimportant}[Newton's first law]
  If all three components of the total force on an object are
  zero, then it will continue in the same state of motion.
\end{lessimportant}

\begin{lessimportant}[Newton's second law]
The components of an object's acceleration are predicted by the equations
\begin{align*}
        a_x  &=  F_{x,total}/m\eqquad,  \\
        a_y  &=  F_{y,total}/m\eqquad, \qquad \text{and}  \\
        a_z  &=  F_{z,total}/m\eqquad.
\end{align*}
\end{lessimportant}

\begin{lessimportant}[Newton's third law]
If two objects A and B interact via forces, then the
components of their forces on each other are equal and opposite:
\begin{align*}
        F_{\text{A on B},x}  &=  -F_{\text{B on A},x}\eqquad,  \\
        F_{\text{A on B},y}  &=  -F_{\text{B on A},y}\eqquad, \qquad \text{and}  \\
        F_{\text{A on B},z}  &=  -F_{\text{B on A},z}\eqquad.
\end{align*}
\end{lessimportant}

<% marg(0) %>
<%
  fig(
    'eg-two-forces-on-object',
    %q{Example \ref{eg:two-forces-on-object}.}
  )
%>
<% end_marg %>
\begin{eg}{Forces in perpendicular directions on the same object}\label{eg:two-forces-on-object}
\egquestion An object is initially at rest. Two constant
forces begin acting on it, and continue acting on it for a
while. As suggested by the two arrows, the forces are
perpendicular, and the rightward force is stronger. What happens?

\eganswer
 Aristotle believed, and many students still do, that
only one force can ``give orders'' to an object at one time.
They therefore think that the object will begin speeding up
and moving in the direction of the stronger force. In fact
the object will move along a diagonal. In the example shown
in the figure, the object will respond to the large
rightward force with a large acceleration component to the
right, and the small upward force will give it a small
acceleration component upward. The stronger force does not
overwhelm the weaker force, or have any effect on the upward
motion at all. The force components simply add together:
\begin{align*}
  F_{x,total} &= F_{1,x} + \cancelto{0}{F_{2,x}} \\
  F_{y,total} &= \cancelto{0}{F_{1,y}} + F_{2,y} 
\end{align*}
\end{eg}

\startdq

\begin{dq}
The figure shows two trajectories, made by splicing together
lines and circular arcs, which are unphysical for an object
that is only being acted on by gravity. Prove that they are
impossible based on Newton's laws.
\end{dq}
<% raw_fig('dq-unphysical-trajectories') %>

<% end_sec() %>\begin{summary}

\begin{vocab}

\vocabitem{component}{the part of a velocity, acceleration, or force
that would be perceptible to an observer who could only see
the universe projected along a certain one-dimensional axis}

\vocabitem{parabola}{the mathematical curve whose graph has $y$
proportional to $x^2$}

\end{vocab}

\begin{notation}

\notationitem{$x$, $y$, $z$}{an object's positions along the $x$, $y$, and $z$ axes}

\notationitem{$v_x$, $v_y$, $v_z$}{the $x$, $y$, and $z$  components of an
object's velocity; the rates of change of the object's
$x$, $y$, and $z$ coordinates}

\notationitem{$a_x$, $a_y$, $a_z$}{the $x$, $y$, and $z$  components of an
object's acceleration; the rates of change of $v_x$, $v_y$, and $v_z$}

\end{notation}

\begin{summarytext}

A force does not produce any effect on the motion of an
object in a perpendicular direction. The most important
application of this principle is that the horizontal motion
of a projectile has zero acceleration, while the vertical
motion has an acceleration equal to $g$. That is, an
object's horizontal and vertical motions are independent.
The arc of a projectile is a parabola.

Motion in three dimensions is measured using three
coordinates, $x$, $y$, and $z$. Each of these coordinates has
its own corresponding velocity and acceleration. We say that
the velocity and acceleration both have $x$, $y$, and $z$ components

Newton's second law is readily extended to three dimensions
by rewriting it as three equations predicting the three
components of the acceleration,
\begin{align*}
a_x=F_{x,total}/m\eqquad, \\
a_y=F_{y,total}/m\eqquad, \\
a_z=F_{z,total}/m\eqquad,
\end{align*}
and likewise for the first and third laws.

\end{summarytext}

\end{summary}
