<% begin_sec("Rigid-body rotation",4,'rigid-body-rotation') %>
<% begin_sec("Kinematics") %>
When a rigid object rotates, every part of it (every atom)
moves in a circle, covering the same angle in the same
amount of time, \figref{toparcs}. Every atom has a different velocity
vector, \figref{topv}. Since all the velocities are different, we
can't measure the speed of rotation of the top by giving a
single velocity. We can, however, specify its speed of
rotation consistently in terms of angle per unit time. Let
the position of some reference point on the top be denoted
by its angle $\theta$, measured in a circle around the axis. For
reasons that will become more apparent shortly, we measure
all our angles in radians. Then the change in the angular
position of any point on the top can be written as $\der\theta$, and
all parts of the top have the same value of $\der\theta$ over a
certain time interval $\der t$. We define the angular velocity, $\omega$\index{angular velocity}
(Greek omega),
\begin{multline*}
        \omega        =          \frac{\der\theta}{\der t}  \qquad m4_ifelse(__sn,1,[:,:],[:.:])        \\
        \text{[definition of
        angular velocity; $\theta$ in units of radians]}
\end{multline*}
m4_ifelse(__sn,1,[:which is similar to, but not the same as, the quantity $\omega$ we defined
earlier to describe vibrations.:],[:%:])
The relationship between $\omega$ and $t$ is
exactly analogous to that between $x$ and $t$ for the motion of
a particle through space.

<% self_check('thetazero',<<-'SELF_CHECK'
If two different people chose two different reference points
on the top in order to define $\\theta\\zu{=0}$, how would their $\\theta$-$t$ graphs
differ? What effect would this have on the angular
velocities?
  SELF_CHECK
  ) %>

<% marg(140) %>
<%
  fig(
    'toparcs',
    %q{The two atoms cover the same angle in a given time interval.}
  )
%>
\spacebetweenfigs
<%
  fig(
    'topv',
    %q{Their velocity vectors, however, differ in both magnitude and direction.}
  )
%>
<% end_marg %>
The angular velocity has units of radians per second, rad/s.
However, radians are not really units at all. The radian
measure of an angle is defined, as the length of the
circular arc it makes, divided by the radius of the circle.
Dividing one length by another gives a unitless quantity, so
anything with units of radians is really unitless. We can
therefore simplify the units of angular velocity, and call
them inverse seconds, $\sunit^{-1}$.

\begin{eg}{A 78-rpm record}
\egquestion
In the early 20th century, the standard format for
music recordings was a plastic disk that held a single song
and rotated at 78 rpm (revolutions per minute). What was the
angular velocity of such a disk?

\eganswer
If we measure angles in units of revolutions and
time in units of minutes, then 78 rpm is the angular
velocity. Using standard physics units of radians/second,
however, we have
\begin{equation*}
                \frac{78\ \zu{revolutions}}{1\ \zu{minute}}
                \times\frac{2\pi\ \zu{radians}}{1\ \zu{revolution}}
                \times\frac{1\ \zu{minute}}{60\ \zu{seconds}}
                         =  8.2\ \sunit^{-1}\eqquad.
\end{equation*}
\end{eg}

In the absence of any torque, a rigid body will rotate
indefinitely with the same angular velocity. If the angular
velocity is changing because of a torque, we define an
angular acceleration,\index{angular acceleration}
\begin{equation*}
        \alpha        =          \frac{\der\omega}{\der t}\eqquad, \qquad        
        \text{[definition of
        angular acceleration]}
\end{equation*}
The symbol is the Greek letter alpha. The units of this
quantity are $\zu{rad}/\zu{s}^2$, or simply $\sunit^{-2}$.
<% marg(0) %>
<%
  fig(
    'analogieskin',
    %q{Analogies between rotational and linear quantities.}
  )
%>
<% end_marg %>

The mathematical relationship between $\omega$ and $\theta$ is the same as
the one between $v$ and $x$, and similarly for $\alpha$ and $a$. We can
thus make a system of analogies, \figref{analogieskin}, and recycle all
the familiar kinematic equations for constant-acceleration
motion.

\begin{eg}{The synodic period}
Mars takes nearly twice as long as the Earth to complete an orbit.
If the two planets are alongside one another on a certain day, then one year
later, Earth will be back at the same place, but Mars will have moved
on, and it will take more time for Earth to finish catching up. Angular
velocities add and subtract, just as velocity
vectors do. If the two planets' angular velocities are $\omega_1$
and $\omega_2$, then the angular velocity of one relative to the other
is $\omega_1-\omega_2$. The corresponding period, $1/(1/T_1-1/T_2)$ is
known as the synodic period.
\end{eg}

\begin{eg}{A neutron star}
\egquestion
A neutron star is initially observed to be
rotating with an angular velocity of $ 2.0\ \sunit^{-1}$, determined via
the radio pulses it emits. If its angular acceleration is a
constant  $- 1.0\times10^{-8}\ \sunit^{-2}$, how many rotations will it
complete before it stops? (In reality, the angular
acceleration is not always constant; sudden changes often
occur, and are referred to as ``starquakes!'')

\eganswer
The equation $v_{f}^2- v_{i}^2\zu{=2} a\Delta x$
 can be translated into
$\omega_{f}^2-\omega_i^2\zu{=2}\alpha\Delta\theta$, giving
\begin{align*}
        \Delta\theta        &=  (\omega_{f}^2-\omega_i^2)/2\alpha \\
                        &        =   2.0\times10^8\ \zu{radians} \\
                        &        =   3.2\times10^7\ \zu{rotations}\eqquad.
\end{align*}
\end{eg}

<% end_sec() %>
<% begin_sec("Relations between angular quantities and motion of a point") %>
It is often necessary to be able to relate the angular
quantities to the motion of a particular point on the rotating
object. As we develop these, we will encounter the first
example where the advantages of radians over degrees become
apparent.

<% marg(0) %>
<%
  fig(
    'vrandvt',
    %q{%
      We construct a coordinate system that coincides with 
      the location and motion of the moving point of interest at a certain moment.
    }
  )
%>
<% end_marg %>
The speed at which a point on the object moves depends on
both the object's angular velocity $\omega$ and the point's
distance $r$ from the axis. We
adopt a coordinate system, \figref{vrandvt}, with an inward (radial) axis
and a tangential axis. The length of the infinitesimal circular arc $\der s$
traveled by the point in a time interval $\der t$ is related
to $\der\theta$ by the definition of radian measure, $\der\theta=\der s/r$, where
positive and negative values of $\der s$ represent the two possible
directions of motion along the tangential axis. We then have
$v_t = \der s/\der t = r\der\theta/\der t = \omega r$, or
\begin{multline*}
        v_t        =  \omega r\eqquad. \hfill
        \shoveright{\text{[tangential velocity of a point at a}}\\
         \text{distance $r$ from the axis of rotation]}
\end{multline*}
The radial component is zero, since the point is not moving
inward or outward,
\begin{multline*}
                v_r        =  0\eqquad.        \hfill
        \shoveright{\text{[radial velocity of a point at a}}\\
        \text{distance $r$ from the axis of rotation]}
\end{multline*}

Note that we had to use the definition of radian measure in
this derivation. Suppose instead we had used units of
degrees for our angles and degrees per second for angular
velocities. The relationship between $\der\theta_{degrees}$ and $\der s$ is
$\der\theta_{degrees} = (360/2\pi)s/r$, where the extra conversion factor
of $(360/2\pi)$ comes from that fact that there are 360
degrees in a full circle, which is equivalent to $2\pi$
radians. The equation for $v_t$ would then have been $v_t =
(2\pi/360)(\omega_{degrees\ per\ second})(r)$, which would have been much
messier. Simplicity, then, is the reason for using radians
rather than degrees; by using radians we avoid infecting all
our equations with annoying conversion factors.

<% marg(0) %>
<%
  fig(
    'aronly',
    %q{%
      Even if the rotating object has zero angular acceleration, 
      every point on it has an acceleration towards the center.
    }
  )
%>
<% end_marg %>
Since the velocity of a point on the object is directly
proportional to the angular velocity, you might expect that
its acceleration would be directly proportional to the
angular acceleration. This is not true, however. Even if the
angular acceleration is zero, i.e., if the object is rotating
at constant angular velocity, every point on it will have an
acceleration vector directed toward the axis, \figref{aronly}.  As derived
on page \pageref{eg:circularaccel}, the magnitude of this
acceleration is
\begin{multline*}
        a_r        =  \omega^2 r\eqquad.        \hfill \shoveright{\text{[radial acceleration of a point}}\\
                        \text{at a distance $r$ from the axis]}
\end{multline*}

For the tangential component, any change in the angular
velocity $\der\omega$ will lead to a change $\der\omega\cdot r$ in the tangential
velocity, so it is easily shown that
\begin{multline*}
        a_t        =  \alpha r\eqquad. \hfill \shoveright{\text{[tangential acceleration of a point}}\\
                \text{at a distance $r$ from the axis]}
\end{multline*}

<% self_check('omegasigns',<<-'SELF_CHECK'
Positive and negative signs of $\\omega$ represent rotation in
opposite directions. Why does it therefore make sense
physically that $\\omega$ is raised to the first power in the
equation for $v_t$ and to the second power in the one for $a_r$?
  SELF_CHECK
  ) %>

\begin{eg}{Radial acceleration at the surface of the Earth}
\egquestion
What is your radial acceleration due to the
rotation of the earth if you are at the equator?

\eganswer
At the equator, your distance from the Earth's
rotation axis is the same as the radius of the spherical
Earth, $ 6.4\times10^6\ \munit$. Your angular velocity
is
\begin{align*}
                \omega &= \frac{2\pi\ \zu{radians}}{1\ \zu{day}} \\
                        &=  7.3\times10^{-5}\ \sunit^{-1}\eqquad,
\end{align*}
which gives an acceleration of
\begin{align*}
        a_{r}        &=  \omega^2 r \\
        &=   0.034\ \munit/\sunit^2\eqquad.
\end{align*}

The angular velocity was a very small number, but the radius
was a very big number. Squaring a very small number,
however, gives a very very small number, so the $\omega^2$ factor
``wins,'' and the final result is small.

If you're
standing on a bathroom scale, this small acceleration is
provided by the imbalance between the downward force of
gravity and the slightly weaker upward normal force of the
scale on your foot. The scale reading is therefore a little
lower than it should be.
\end{eg}

<% end_sec() %>
<% begin_sec("Dynamics") %>
If we want to connect all this kinematics to anything
dynamical, we need to see how it relates to torque and
angular momentum. Our strategy will be to tackle angular
momentum first, since angular momentum relates to motion,
and to use the additive property of angular momentum: the
angular momentum of a system of particles equals the sum of
the angular momenta of all the individual particles. The
angular momentum of one particle within our
rigidly rotating object, $L=mv_\perp r$, can be rewritten as
$L=r\:p\:\sin\:\theta$, where $r$ and $p$ are the
magnitudes of the particle's $\vc{r}$ and momentum vectors, and $\theta$
is the angle between these two vectors. (The \vc{r} vector points
outward perpendicularly from the axis to the particle's
position in space.) In rigid-body rotation the angle $\theta$ is
90\degunit, so we have simply $L=r p$. Relating this to angular
velocity, we have $L=rp=(r)(mv)=(r)(m\omega r)=mr^2\omega$. The particle's
contribution to the total angular momentum is proportional
to $\omega$, with a proportionality constant $mr^2$. We refer to $mr^2$
as the particle's contribution to the object's total \emph{moment
of inertia},\index{moment of inertia}
 $I$, where ``moment'' is used in the sense of
``important,'' as in ``momentous''  --- a bigger value of $I$
tells us the particle is more important for determining the
total angular momentum. The total moment of inertia
is
\begin{multline*}
        I        =          \sum{m_i r_i^2}\eqquad,
                 \hfill \shoveright{\text{[definition of the moment of inertia;}} \\
                \hfill \shoveright{\text{for rigid-body
                rotation in a plane; $r$ is the distance}}\\
                \hfill \shoveright{\text{from the axis, measured perpendicular
                to the axis]}}
\end{multline*}
The angular momentum of a rigidly rotating body is then
\begin{multline*}
        L        =  I \omega\eqquad. \hfill
                        \shoveright{\text{[angular momentum of}}\\
        \text{rigid-body rotation in a plane]}
\end{multline*}

Since torque is defined as $\der L/\der t$, and a rigid body has
a constant moment of inertia, we have $\tau=\der L/\der t
=I\der\omega/\der t=I\alpha$,
\begin{multline*}
        \tau        =  I \alpha\eqquad,        \hfill
                \shoveright{\text{[relationship between torque and}}\\
        \text{angular acceleration for rigid-body rotation in a plane]}
\end{multline*}
which is analogous to $F=ma$.

<% marg(100) %>
<%
  fig(
    'analogiesdyn',
    %q{Analogies between rotational and linear quantities.}
  )
%>
<% end_marg %>

The complete system of analogies between linear motion and
rigid-body rotation is given in figure \figref{analogiesdyn}.

\begin{eg}{A barbell}\label{eg:barbell}
\egquestion
The barbell shown in figure \figref{barbell} consists of two
small, dense, massive balls at the ends of a very light rod.
The balls have masses of 2.0 kg and 1.0 kg, and the length
of the rod is 3.0 m. Find the moment of inertia of the rod
(1) for rotation about its center of mass, and (2) for
rotation about the center of the more massive ball.

<% marg(30) %>
<%
  fig(
    'barbell',
    %q{Example \ref{eg:barbell}}
  )
%>
<% end_marg %>

\eganswer
(1) The ball's center of mass lies 1/3 of the way from the
greater mass to the lesser mass, i.e., 1.0 m from one and 2.0
m from the other. Since the balls are small, we approximate
them as if they were two pointlike particles. The moment of
inertia is
\begin{align*}
         I        &=  \zu{(2.0\ kg)(1.0\ m)}^2 + \zu{(1.0\ kg)(2.0\ m)}^2 \\
                &=   2.0\ \kgunit\unitdot\munit^2
                         +  4.0\ \kgunit\unitdot\munit^2  \\
                &=   6.0\ \kgunit\unitdot\munit^2
\end{align*}
Perhaps counterintuitively, the less massive ball
contributes far more to the moment of inertia.

(2) The big ball theoretically contributes a little bit to
the moment of inertia, since essentially none of its atoms
are exactly at $r$=0. However, since the balls are said to be
small and dense, we assume all the big ball's atoms are so
close to the axis that we can ignore their small
contributions to the total moment of inertia:
\begin{align*}
         I                &=  \zu{(1.0\ kg)(3.0\ m)}^2 \\
                &        =   9.0\ \kgunit\unitdot\munit^2
\end{align*}
This example shows that the moment of inertia depends on the
choice of axis. For example, it is easier to wiggle a pen
about its center than about one end.
\end{eg}

\begin{eg}{The parallel axis theorem}
\egquestion
Generalizing the previous example, suppose we pick
any axis parallel to axis 1, but offset from it by a
distance $h$. Part (2) of the previous example then
corresponds to the special case of $h=- 1.0\ \munit$ (negative being
to the left). What is the moment of inertia about this new
axis?

\eganswer
The big ball's distance from the new axis is 
$\zu{(1.0\ m)+} h$, and the small one's is $\zu{(2.0\ m)-} h$. The new moment of
inertia is
\begin{align*}
         I        &=  \zu{(2.0 kg)}[\zu{(1.0 m)+} h]^2 + \zu{(1.0 kg)}[\zu{(2.0 m)}- h]^2 \\
                &        =   6.0 \ \kgunit\unitdot\munit^2         
                        + \zu{(4.0}\ \kgunit\unitdot\munit) h
                        - \zu{(4.0}\ \kgunit\unitdot\munit) h
                        + \zu{(3.0 kg)} h^2\eqquad.
\end{align*}
The constant term is the same as the moment of inertia about
the center-of-mass axis, the first-order terms cancel out,
and the third term is just the total mass multiplied by $h^2$.
The interested reader will have no difficulty in
generalizing this to any set of particles (problem \ref{hw:parallel-axis-theorem}, p.~\pageref{hw:parallel-axis-theorem}), resulting in the
parallel axis theorem:\index{parallel axis theorem}
 If an object of total mass $M$ rotates
about a line at a distance $h$ from its center of mass, then
its moment of inertia equals $I_{cm}+ Mh^2$, where $I_{cm}$ is the
moment of inertia for rotation about a parallel line through
the center of mass.
\end{eg}

\begin{eg}{Scaling of the moment of inertia}
\egquestion
(1) Suppose two objects have the same mass and the
same shape, but one is less dense, and larger by a factor $k$.
How do their moments of inertia compare? \\
(2) What if the
densities are equal rather than the masses?

\eganswer
(1) This is like increasing all the distances
between atoms by a factor $k$. All the $r$'s become greater by
this factor, so the moment of inertia is increased by a
factor of $k^2$.\\
(2) This introduces an increase in mass by a
factor of $k^3$, so the moment of inertia of the bigger object
is greater by a factor of $k^5$.
\end{eg}

<% end_sec() %>
<% begin_sec("Iterated integrals",nil,'iterated-int') %>
In various places in this book, starting with subsection \ref{subsec:moi-integ},
we'll come across integrals stuck inside other integrals. These are known as
iterated integrals, or double integrals, triple integrals, etc. Similar concepts
crop up all the time even when you're not doing calculus, so let's start by
imagining such an example. Suppose you want to count how many squares there are
on a chess board, and you don't know how to multiply eight times eight. You
could start from the upper left, count eight squares across, then continue with
the second row, and so on, until you how counted every square, giving the result
of 64. In slightly more formal mathematical language, we could write the following
recipe: for each row, $r$, from 1 to 8, consider the columns, $c$, from 1 to 8,
and add one to the count for each one of them. Using the sigma notation, this
becomes
\begin{equation*}
  \sum_{r=1}^8 \sum_{c=1}^8 1\eqquad.
\end{equation*}
If you're familiar with computer programming, then you can think of this as
a sum that could be calculated using a loop nested inside another loop.
To evaluate the result (again, assuming we don't know how to multiply, so we
have to use brute force), we can first evaluate the inside sum, which equals
8, giving
\begin{equation*}
  \sum_{r=1}^8 8\eqquad.
\end{equation*}
Notice how the ``dummy'' variable $c$ has disappeared. Finally we do the outside
sum, over $r$, and find the result of 64.

Now imagine doing the same thing with the pixels on a TV screen. The electron
beam sweeps across the screen, painting the pixels in each row, one at a time.
This is really no different than the example of the chess board, but because the
pixels are so small, you normally think of the image on a TV screen as continuous
rather than discrete.
This is the idea of an integral in calculus.
Suppose we want to find the area of a rectangle of width
$a$ and height $b$, and we don't know that we can just multiply to get the
area $ab$. The brute force way to do this is to break up the rectangle into
a grid of infinitesimally small squares, each having width $\der x$ and
height $\der y$, and therefore the infinitesimal area $\der A = \der x \der y$.
For convenience, we'll imagine that the rectangle's
lower left corner is at the origin. Then the area is given by this integral:
\begin{align*}
  \text{area} &= \int_{y=0}^b \int_{x=0}^a \der A \\
              &= \int_{y=0}^b \int_{x=0}^a \der x \der y 
\end{align*}
Notice how the leftmost integral sign, over $y$, and the rightmost differential, $\der y$,
act like bookends, or the pieces of bread on a sandwich. Inside them, we have
the integral sign that runs over $x$, and the differential $\der x$ that matches
it on the right. Finally, on the innermost layer, we'd normally
have the thing we're integrating, but here's it's 1, so I've omitted it. Writing the
lower limits of the integrals with $x=$ and $y=$ helps to keep it straight which
integral goes with which differential. The result is
\begin{align*}
  \text{area} &= \int_{y=0}^b \int_{x=0}^a \der A \\
              &= \int_{y=0}^b \int_{x=0}^a \der x \der y \\
              &= \int_{y=0}^b \left(\int_{x=0}^a \der x\right) \der y \\
              &= \int_{y=0}^b a \der y \\
              &= a \int_{y=0}^b \der y \\
              &= ab\eqquad.
\end{align*}

\begin{eg}{Area of a triangle}
\egquestion Find the area of a 45-45-90 right triangle having legs $a$.

\eganswer Let the triangle's hypotenuse run from the origin to the point $(a,a)$,
and let its legs run from the origin to $(0,a)$, and then to $(a,a)$. In other
words, the triangle sits on top of its hypotenuse. Then the integral can be set
up the same way as the one before, but for a particular value of $y$, values of
$x$ only run from 0 (on the $y$ axis) to $y$ (on the hypotenuse). We then have
\begin{align*}
  \text{area} &= \int_{y=0}^a \int_{x=0}^y \der A \\
              &= \int_{y=0}^a \int_{x=0}^y \der x \der y \\
              &= \int_{y=0}^a \left(\int_{x=0}^y \der x\right) \der y \\
              &= \int_{y=0}^a y \der y \\
              &= \frac{1}{2}a^2
\end{align*}
Note that in this example, because the upper end of the $x$ values depends
on the value of $y$, it makes a difference which order we do the integrals
in. The $x$ integral has to be on the inside, and we have to do it first.
\end{eg}

\begin{eg}{Volume of a cube}
\egquestion Find the volume of a cube with sides of length $a$.

\eganswer This is a three-dimensional example, so we'll have integrals
nested three deep, and the thing we're integrating is the volume
$\der V = \der x \der y \der z$.

\begin{align*}
  \text{volume} &= \int_{z=0}^a \int_{y=0}^a \int_{x=0}^a \der x \der y \der z \\
              &= \int_{z=0}^a \int_{y=0}^a a \der y \der z \\
              &= a \int_{z=0}^a \int_{y=0}^a \der y \der z \\
              &= a \int_{z=0}^a a \der z \\
              &= a^3
\end{align*}
\end{eg}

\begin{eg}{Area of a circle}
\egquestion Find the area of a circle.

\eganswer To make it easy, let's find the area of a semicircle and then double it.
Let the circle's radius be $r$, and let it be centered on the origin and bounded
below by the $x$ axis. Then the curved edge is given by the equation $r^2=x^2+y^2$,
or $y=\sqrt{r^2-x^2}$. Since the $y$ integral's limit depends on $x$, the $x$
integral has to be on the outside.
The area is
\begin{align*}
  \text{area} &= \int_{x=-r}^r \int_{y=0}^{\sqrt{r^2-x^2}} \der y \der x\\
              &= \int_{x=-r}^r \sqrt{r^2-x^2} \der x\\
              &= r \int_{x=-r}^r \sqrt{1-(x/r)^2} \der x\eqquad.
\intertext{Substituting $u=x/r$,}
  \text{area} & = r^2 \int_{u=-1}^1 \sqrt{1-u^2} \der u \\
\end{align*}
The definite integral equals $\pi$, as you can
find using a trig substitution or simply by looking it
up in a table, and the result is, as expected, $\pi r^2/2$ for the area of
the semicircle.
\end{eg}

 % 
<% end_sec() %>
<% begin_sec("Finding moments of inertia by integration",nil,'moi-integ') %>
When calculating the moment of inertia of an ordinary-sized
object with perhaps $10^{26}$ atoms, it would be impossible to do
an actual sum over atoms, even with the world's fastest
supercomputer. Calculus, however, offers a tool, the
integral, for breaking a sum down to infinitely many small
parts. If we don't worry about the existence of atoms, then
we can use an integral to compute a moment of inertia as if
the object was smooth and continuous throughout, rather than
granular at the atomic level. Of course this granularity
typically has a negligible effect on the result unless the
object is itself an individual molecule. This subsection consists
of three examples of how to do such a computation, at three
distinct levels of mathematical complication.

<% begin_sec("Moment of inertia of a thin rod") %>
What is the moment of inertia of a thin rod of
mass $M$ and length $L$ about a line perpendicular to the rod
and passing through its center?
We generalize the discrete sum
\begin{align*}
        I        &=          \sum{m_i r_i^2}\\
\intertext{to a continuous one,}
        I        &= \int r^2 \der m \\
                &= \int_{-L/2}^{L/2} x^2\:\frac{M}{L}\der x \qquad \text{[$r=|x|$, so $r^2=x^2$]} \\
                &= \frac{1}{12}ML^2
\end{align*}

In this example the object was one-dimensional, which made
the math simple. The next example shows a strategy that can
be used to simplify the math for objects that are
three-dimensional, but possess some kind of symmetry.

<% end_sec() %>
<% begin_sec("Moment of inertia of a disk") %>
What is the moment of inertia of a disk of radius
$b$, thickness $t$, and mass $M$, for rotation about its central
axis?

We break the disk down into concentric circular
rings  of thickness $\der r$. Since all the mass in a given
circular slice has essentially the same value of $r$ (ranging
only from $r$ to $r+\der r$), the slice's contribution to the total
moment of inertia is simply $r^2\der m$. We then have
\begin{align*}
        I        &= \int r^2 \der m \\
                &= \int r^2 \rho\der V\eqquad,
\end{align*}
where $V=\pi b^2 t$ is the total volume, $\rho=M/V=M/\pi b^2 t$ is the
density, and the volume of one
slice can be calculated as the volume enclosed by its outer
surface minus the volume enclosed by its inner surface,
$\der V= \pi (r+\der r)^2 t - \pi r^2 t = 2\pi tr \der r$.
\begin{align*}
        I        &=  \int_0^b r^2 \frac{M}{\pi b^2 t}\:2\pi t\:r\der r\\
                &=  \frac{1}{2}Mb^2\eqquad.
\end{align*}

In the most general case where there is no symmetry about
the rotation axis, we must use iterated integrals, as discussed
in subsection \ref{subsec:iterated-int}. The
example of the disk possessed two types of symmetry with
respect to the rotation axis: (1) the disk is the same when
rotated through any angle about the axis, and (2) all slices
perpendicular to the axis are the same. These two symmetries
reduced the number of layers of integrals from three to one.
The following example possesses only one symmetry, of type
(2), and we simply set it up as a triple integral. You may not
have seen multiple integrals yet in a math course. If so, just
skim this example.

<% end_sec() %>
<% begin_sec("Moment of inertia of a cube") %>
What is the moment of inertia of a cube of side $b$,
for rotation about an axis that passes through its center
and is parallel to four of its faces?
Let the origin be at the center of the cube, and
let $x$ be the rotation axis.
\begin{align*}
        I        &=  \int r^2 \der m \\
                &= \rho \int r^2 \der V \\
                &= \rho \int_{-b/2}^{b/2} \int_{-b/2}^{b/2} \int_{-b/2}^{b/2} \left(y^2+z^2\right)
                                         \der x\der y\der z \\
                &= \rho b \int_{-b/2}^{b/2} \int_{-b/2}^{b/2}  \left(y^2+z^2\right)
                                         \der y \der z 
\end{align*}
The fact that the last step is a trivial integral results
from the symmetry of the problem. The integrand of the
remaining double integral breaks down into two terms, each
of which depends on only one of the variables, so we break
it into two integrals,
\begin{equation*}
                I = \rho b \int_{-b/2}^{b/2} \int_{-b/2}^{b/2}  y^2 \der y\der z
                        + \rho b \int_{-b/2}^{b/2} \int_{-b/2}^{b/2}  z^2 \der y\der z
\end{equation*}
which we know have identical results. We therefore only need
to evaluate one of them and double the result:
\begin{align*}
        I        &= 2\rho b \int_{-b/2}^{b/2} \int_{-b/2}^{b/2}  z^2 \der y\der\: z \\
                &= 2 \rho b^2 \int_{-b/2}^{b/2} z^2 \der z \\
                &= \frac{1}{6} \rho b^5 \\
                &= \frac{1}{6} M b^2
\end{align*}

Figure \figref{moments-of-inertia} shows the moments of inertia of some
shapes, which were evaluated with techniques like these.\index{moment of inertia!tabulated for various shapes}

<%
  fig(
    'moments-of-inertia',
    %q{Moments of inertia of some geometric shapes.},
    {
      'width'=>'wide',
      'sidecaption'=>true
    }
  )
%>

\begin{eg}{The hammer throw}
\egquestion In the men's Olympic hammer throw, a steel ball of radius 6.1 cm is swung on the
end of a wire of length 1.22 m. What fraction of the ball's angular momentum
comes from its rotation, as opposed to its motion through space?

\eganswer It's always important to solve problems symbolically first, and plug in numbers
only at the end, so let the radius of the ball be $b$, and the length of the wire $\ell$.
If the time the ball takes to go once around the circle is $T$, then
this is also the time it takes to revolve once around its own axis. Its speed
is $v=2\pi\ell/T$, so its angular momentum due to its motion through space
is $mv\ell=2\pi m\ell^2/T$. Its angular momentum due to its rotation around its
own center is $(4\pi/5)mb^2/T$. The ratio of these two angular momenta is
$(2/5)(b/\ell)^2=1.0\times10^{-3}$. The angular momentum due to the ball's
spin is extremely small.
\end{eg}

\begin{eg}{Toppling a rod}\label{eg:toppling-rod}
\egquestion A rod of length $b$ and mass $m$ stands upright. We want
to strike the rod at the bottom, causing it to fall and land flat.
Find the momentum, $p$, that should be delivered, in terms of $m$, $b$,
and $g$. Can this really be done without having the rod scrape on the floor?
<% marg(0) %>
<% fig(
    'eg-toppling-rod',
    'Example \ref{eg:toppling-rod}.'
  )
%>
<% end_marg %>

\eganswer This is a nice example of a question that can very nearly be
answered based only on units. Since the three variables, $m$, $b$,
and $g$, all have different units, they can't be added or subtracted.
The only way to combine them mathematically is by multiplication or division.
Multiplying one of them by itself is exponentiation, so in general
we expect that the answer must be of the form
\begin{equation*}
  p = A m^j b^k g^l\eqquad,
\end{equation*}
where $A$, $j$, $k$, and $l$ are unitless constants. The result has
to have units of $\kgunit\unitdot\munit/\sunit$. To get kilograms to
the first power, we need
\begin{equation*}
  j=1\eqquad,
\end{equation*}
meters to the first power requires
\begin{equation*}
  k+l=1\eqquad,
\end{equation*}
and
seconds to the power $-1$ implies
\begin{equation*}
  l=1/2\eqquad.
\end{equation*}
We find $j=1$, $k=1/2$, and $l=1/2$, so the solution must be of the form
\begin{equation*}
  p = A m\sqrt{bg}\eqquad.
\end{equation*}
Note that no physics was required!

Consideration of units, however, won't help us to find the unitless constant
$A$. Let $t$ be the time the rod takes to fall, so that $(1/2)gt^2=b/2$.
If the rod is going to land exactly on its side, then the number of revolutions
it completes while in the air must be 1/4, or 3/4, or 5/4,  \ldots, but all the
possibilities greater than 1/4 would cause the head of the rod to collide with
the floor prematurely. The rod must therefore rotate at a rate that would
cause it to complete a full rotation in a time $T=4t$, and it has angular
momentum $L=(\pi/6)mb^2/T$.

The momentum lost by the object striking
the rod is $p$, and by conservation of momentum, this is the amount of
momentum, in the horizontal direction, that the rod acquires. In other words,
the rod will fly forward a little. However, this has no effect on the solution
to the problem. More importantly, the object striking the rod loses angular
momentum $bp/2$, which is also transferred to the rod. Equating this to the
expression above for $L$, we find $p=(\pi/12)m\sqrt{bg}$.

Finally, we need to know whether this can really be done without having the
foot of the rod scrape on the floor. The figure shows that the answer is no
for this rod of finite width, but it appears that the answer would be yes for
a sufficiently thin rod. This is analyzed further in homework problem \ref{hw:toppling-rod}
on page \pageref{hw:toppling-rod}.

\end{eg}

\vfill

<% end_sec() %>
<% end_sec() %>
<% end_sec('rigid-body-rotation') %>
