First, let's account for every object that's touching her: the floor and the roof.
Any time two solid objects are in contact, we expect a normal force, which is the
force that keeps them from passing through each other. Normal forces are repulsive,
which means here that the roof's force on her head is down (away from itself),
and the floor's force on her feet is up (away from itself).
There could also be frictional
forces, but in this problem there is symmetry between left and right, so it wouldn't
make sense for frictional forces to exist here --- if they did, there would be no
way to decide which way they should point.

In addition to these contact forces, we will have a non-contact force: the earth's
gravity.

The physical reasoning above establishes the left-hand column of the table below.
Once we've established the left-hand column, the right-hand column can be generated
purely by manipulating the words and symbols, without further recourse to physical
insight. By Newton's third law, we interchange the two objects and reverse the arrow.
The type of the force is the same.

\begin{tabular}{|p{70mm}|p{70mm}|}
\hline
\emph{force acting on woman}  &   \emph{force related to it by Newton's third law} \\
\hline
roof's normal force on woman, $\downarrow$ & woman's normal force on roof, $\uparrow$\\
\hline
floor's normal force on woman, $\uparrow$ & woman's normal force on floor, $\downarrow$\\
\hline
planet earth's gravitational force on woman, $\downarrow$  &  woman's gravitational force on earth, $\uparrow$\\
\hline
\end{tabular}
